### 2015-04-17 implementation-concretization.tex
Lots of edits.

>>>>

Before it builds a package, Spack ensures the following conditions:
\newline

\begin{tabular}{l}
(1) All dependencies are present in the DAG. \\
(2) Parameters are set on all packages in the DAG. \\
\end{tabular}\newline

\noindent
The install process then constructs a package object for each node in the spec DAG
and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
{\tt install()} method, using a sub-DAG rooted at the package to be installed as the {\tt spec}
parameter to {\tt install()}. Package maintainers are responsible for examining the spec's
configuration and adjusting the build if necessary.

In our experience, much complexity in build scripts arises from the need for the script to
query the environment and determine how to build.  In spack, build scripts are free from this 
concern, because the package author is guaranteed that a package's build spec is concrete
by the time {\tt install()} is called. There is thus no need to perform complicated state checks,
just to query the spec for its configuration.

### 2015-04-17 implementation-packages.tex
Fix typos.  Add gperftools section.
Lots of edits.

%\subsubsection{Package Files}
\begin{figure}
	\includegraphics[width=\columnwidth]{code/mpileaks.pdf}
	\caption{
		Spack package for the {\tt mpileaks} tool.
		\label{fig:mpileaks}
	}
\end{figure}

In Spack, packages are build scripts that describe how to build software
artifacts.  Each package is a class written in pure Python to describe
{\it specific} build instructions, and it must extend the {\tt Package}
base class, which implements the more general parts of the build process.
Spack implements a simple, embedded domain-specific language (DSL) to make
packaging easier; it adds special directives such as {\tt depends\_on} and
{\tt version} that add metadata to the class.

Figure~\ref{fig:mpileaks} shows the package for \mpileaks, an LLNL-developed
tool for finding leaks in MPI programs.
Inside the {\tt MpiLeaks} class, the package provides a text description
and a homepage, as well as 
a download URL.  Next, two {\tt version} directives identify known versions
of the package, and MD5 checksums ensure they can be downloaded safely.
Below this, two {\tt depends\_on}
directives indicate prerequisites that must be installed before \mpileaks.
Last, each package defines an {\tt install()} method, which contains the
commands used to build the package.  Spack's DSL allows shell
commands to be invoked as Python functions. Here, the {\tt install()} 
method invokes the familiar {\tt configure}, {\tt make}, and
{\tt make install} build commands, as a shell script would.

>>>>

%\subsubsection{Package Files}
\begin{figure}
\begin{minted}[linenos,
               numbersep=5pt,
			   fontsize=\scriptsize,
               frame=lines,
               framesep=2mm]{python}
from spack import *

class Mpileaks(Package):
    """Tool to detect and report leaked MPI objects like
       MPI_Requests and MPI_Datatypes."""

    homepage = "https://github.com/hpc/mpileaks"
    url = homepage + "/releases/download/v1.0/mpileaks-1.0.tar.gz"

    version('1.0', '8838c574b39202a57d7c2d68692718aa')
    version('1.1', '4282eddb08ad8d36df15b06d4be38bcb')

    depends_on("mpi")
    depends_on("callpath")

    def install(self, spec, prefix):
        configure("--prefix=" + prefix,
                  "--with-callpath=" + spec['callpath'].prefix)
         make()
         make("install")
\end{minted}
	\caption{
		Spack package for the {\tt mpileaks} tool.
		\label{fig:mpileaks}
	}
\end{figure}

In Spack, packages are Python scripts that build software artifacts.  
Each package is a class that extends a generic {\tt Package}
base class.  {\tt Package} implements the bulk of the build process, but
subclasses provide their own {\tt install()} method to handle the
specifics of particular packages. Additionally,
Spack implements an embedded domain-specific language (DSL).  It provides
special directives such as {\tt depends\_on} and {\tt version} that add
metadata to the package class.

Figure~\ref{fig:mpileaks} shows the package for \mpileaks, a tool developed
at LLNL to find memory leaks in MPI programs.
The {\tt MpiLeaks} class first provides simple metadata: a text description,
a homepage, and a download URL.  
Next, two {\tt version} directives identify known versions and provide
MD5 checksums used to verify downloads.
Below this, two {\tt depends\_on}
directives indicate prerequisite packages that must be installed before \mpileaks.
Last, the {\tt install()} method contains the commands used to build
the tool.  Spack's DSL allows shell commands to be invoked as Python functions,
and the {\tt install()} method invokes {\tt configure},
{\tt make}, and {\tt make install} commands as a shell script would.

### 2015-04-17 implementation-specs.tex
Fix typos.  Add gperftools section.
Lots of edits.

\begin{table*}\centering
\begin{tabular}{|r|p{2.4in}|p{4in}|}
\hline
& {\bf Spec} & {\bf Meaning} \\
\hline
\hline
1&\small\verb|mpileaks|                         & \small {\tt mpileaks} package, no constraints. \\\hline
2&\small\verb|mpileaks@1.1.2|                   & \small {\tt mpileaks} package, version 1.1.2. \\\hline
3&\small\verb|mpileaks@1.1.2 %gcc|              & \small {\tt mpileaks} package, version 1.1.2, built with {\tt gcc} at the default version. \\\hline
4&\small\verb|mpileaks@1.1.2 %intel@14.1 +debug| & \small {\tt mpileaks} package, version 1.1.2, built with Intel compiler version 14.1, \newline with the ``debug'' build option. \\\hline
5&\small\verb|mpileaks@1.1.2 =bgq|              & \small {\tt mpileaks} package, version 1.1.2, built for the Blue Gene/Q platform (BG/Q). \\\hline
6&\small\verb|mpileaks@1.1.2 ^mvapich2@1.9|     & \small {\tt mpileaks} package version 1.1.2, using {\tt mvapich2}  version 1.9 for MPI. \\\hline
7&\small\verb|mpileaks @1.2:1.4 %gcc@4.7.5 -debug =bgq| \newline
      \verb|  ^callpath @1.1 %gcc@4.7.2| \newline
      \verb|  ^openmpi @1.4.7|                & \small%
      {\tt mpileaks} at any version between 1.2 and 1.4 (inclusive), built with gcc 4.7.5, 
      without the debug option, for BG/Q, linked with {\tt callpath} version 1.1
      and building {\tt callpath} with {\tt gcc} version 4.7.2, linked with {\tt openmpi} version 1.4.7.    \\
\hline
\end{tabular}
\caption{
	Example Spack build specs and their meanings.
	\label{tab:specs}
}
\end{table*}

\begin{figure}
	\subfigure[{\tt mpileaks}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
		\label{fig:specs-mpileaks}
	}
	\subfigure[{\tt mpileaks \^{}callpath\@1.0+debug \^{}libelf\@0.8.11}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
		\label{fig:specs-mpileaks-abstract}
	}
	\subfigure[Concrete version of (b)]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
		\label{fig:specs-mpileaks-concrete}
	}
	\caption{
		Specs for the {\tt mpileaks} package.
		\label{fig:specs}
	}
\end{figure}

@@

The package shown in Figure~\ref{fig:mpileaks} seems simple, but even this short script
allows Spack to build many versions of the {\tt mpileaks} package.  This is because builds
in Spack are parameterized.  In traditional port systems, code is structured to build a single
version of a package, but in Spack, each package file is a {\it template} that can build 
in many different ways.  Spack's representation of a configuration is a build specification, 
or {\tt spec}.

\subsubsection{Spec structure}
To understand how specs works, consider the package structure.  The  {\tt Mpileaks} package
is a class, and its metadata ({\tt version}, {\tt depends\_on}, etc.) statically describe
its relationships with other packages.  In this case, there are two direct dependencies: 
the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
for each of these packages to construct a graph of their relationships.  
To avoid issues with incompatible ABIs, and to guarantee a consistent build, the DAG is constructed
so that there is only ever one version of any particular package.

For {\tt mpileaks}, 
the initial graph is shown in Figure~\ref{fig:specs-mpileaks}.  This graph is the in-memory
representation of a spec.

Each package, or node, in the spec has a number of configuration parameters that control
how the package will be built.  These are the version, compiler to build with, compiler version, 
named build options (or {\it variants} like +debug or -debug), and the architecture to
build for.  When Spack runs to install {\tt mpileaks}, it constructs a package object for each node
in the spec DAG, and it traverses the DAG in a bottom-up fashion.  At each node, it
invokes the package's {\tt install()} method, passing in a sub-DAG rooted at the
package to be installed.  This is the {\tt spec} parameter in Figure~\ref{fig:mpileaks}.
Package maintainers are responsible for examining the spec's configuration and adjusting the 
build if necessary.

\subsubsection{Spec syntax}
A spec DAG has many degrees of freedom, and it is not reasonable to expect users to
understand all of them.  In our experience at LLNL, users typically have a small number of 
build constraints that they care about, and the others only confuse them.  This is part of why
the HPC software ecosystem is so difficult to manage: there are simply too many parameters.
However, the small set of important build constraints can be very specific, so we need the
ability to specify all aspects of the configuration space if necessary.  For this reason, 
we introduce a telescoping syntax for package specs that allows users of Spack to
specify {\it only those constraints they care about}, but to be {\it specific when necessary}.
This is the true power of our syntax---specs are concise enough to use on the command line,
and users can request installs or query existing installs by their structure.

Table~\ref{tab:specs} shows examples of specs, ranging from very simple to very complex.
Next to each is a description of the spec in plain language.  In the most basic case,
shown on line 1 in the table, a user can simply name the package they want to use, and have it
built and installed a default configuration.  On the command line this would take the form:
\newline\newline
{\tt spack install mpileaks}
\newline

If the user wants more specificity, {\tt mpileaks} can be augmented with additional
qualifiers.  Spack install will handle selecting a build configuration that satisfies
the user's request.  We describe additional qualifiers below.

{\bf Versions.}
Versions are specified by supplying {\tt @<version>} after the package name (line 2). 
Versions can be precise (e.g. {\tt @4.5.1}) or 
they can be ambiguous (e.g., @4) or a colon-separated, inclusive range (e.g., @4.4:4.5.1),
and Spack will pick an available version from the specified range.
The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
in our experience users frequently want bleeding-edge versions.  Package managers
frequently lag behind the latest releases. 
Spack has a capability to extrapolate URLs from versions,
using the package's {\tt url} attribute as a model\footnote{This works
for packages with consistently named URLs}.  The user can request a specific
version on the command line, even if it is unknown to Spack,
and Spack will attempt to install it.  Spack also uses the same
model to scrape webpages and find new versions as they become available.

{\bf Compilers.}
To specify the compiler (line 3), the user
simply adds {\tt \%} followed by its name, and if a version {\it of the compiler} 
is desired, the same {\tt @} syntax can be used.  Here, {\tt gcc} refers to the full
gcc compiler suite: C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
compiler toolchains if they are in the user's PATH or they can be registered manually
through a configuration file.

{\bf Variants.}
To handle build-time options like debug flags and optional components, spack
allows named flags, or {\it variants} to be associated with packages.  
In the example (line 4), {\tt mpileaks} can be built with a {\tt debug} option by
adding {\tt +debug}.  The option can be similarly disabled with {\tt -debug}.  
This allows package authors to provide package-specific build options without 
having to extend Spack.  We use names for these to simplify the spec
syntax, and to prevent Spack's configuration space from becoming too fine-grained.
It would violate our goal of conciseness, for example, to include detailed 
compiler flags in the spec syntax, but known sets of flags can be exported using
named variants.

{\bf Cross-compileation.}
To support cross-compilation, Spack includes the platform in the package spec (line 5).
Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bbq}.  They are
specified per-package; this allows front-end tools to depend on their back-end measurement
libraries with a {\it different} architecture on cross-compiled machines.

{\bf Dependencies.}
Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
The user can specify all of the above information not only for the package being
installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
\verb|^| and the dependency's name.  If we need to build a new version with a specific
version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
to the spec, and it will build with that MPI version instead of the default.
This can also be done for multiple libraries in the same spec (line 7).  
As mentioned, Spack guarantees that there is only one version of any package in 
the spec DAG.  So, within the same DAG, each dependency can be uniquely identified by 
only its package name.  This frees the user from having to worry about DAG connectivity;
the user only needs to know that a package depends, somehow, indirectly on {\tt mvapich}
to add this constraint
uniquely identifies a dependency, and dependency constraints can be specified in any
order.

\subsubsection{Constraining dependencies}

@@

\subsubsection{Version extrapolation}

\subsubsection{Querying installed packages}

>>>>

\begin{figure}
	\subfigure[Spec for {\tt mpileaks}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
		\label{fig:specs-mpileaks}
	}
	\subfigure[{\tt mpileaks@2.3}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-version}
		\label{fig:specs-mpileaks-version}
	}
	\subfigure[{\tt mpileaks@2.3 \^{}callpath@1.0+debug \^{}libelf@0.8.11}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
		\label{fig:specs-mpileaks-abstract}
	}
	\subfigure[Spec (b), concretized.]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
		\label{fig:specs-mpileaks-concrete}
	}
	\caption{
		Specs for the {\tt mpileaks} package.
		\label{fig:specs}
	}
\end{figure}

@@

Using the simple script in Figure~\ref{fig:mpileaks}, Spack can build many different
versions and configurations of the {\tt mpileaks} package.  In traditional port systems, 
package code is structured to build a single version of a package, but in Spack, each
package file is a {\it template} that can be built in many different ways.  The build is
{\it parameterized} so that it can be configured many different ways.  
Spack calls a single build configuration a {\it spec},
 and the {\tt spec} object passed to {\tt install()} 
is how the Spack system encapsulates build parameters for package authors.

\subsubsection{Structure}
To understand how specs work, consider the {\tt mpileaks} package structure.  
Metadata in the {\tt Mpileaks} class ({\tt version}, {\tt depends\_on}, etc.) describe
its relationships with other packages.  There are two direct dependencies: 
the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
for each dependency and constructs a graph of their relationships.  The result
is a directed, acyclic graph (DAG) of dependencies\footnote{Spack currently disallows
circular dependencies.}.  
%
To guarantee a consistent build, and to avoid ABI incompatibility, the DAG
is constructed so that there is only {\it one} version of any particular package.  Note
the distinction: while Spack can install arbitrarily many configurations of any package,
no two configurations of the same package will ever appear in the same build DAG.

DAGs for {\tt mpileaks} are shown in Figure~\ref{fig:specs}.
Each node is a package and has five configuration parameters that control
how it will be built: 1) the package version, 2) the compiler to
build with, 3) the compiler version, 4) named compile-time build options, or {\it variants},
and 5) the target architecture.

\subsubsection{Configuration Complexity}
A spec DAG has many degrees of freedom, and it is not reasonable to expect users to
understand or specify all of them.  In our experience at LLNL, the typical user
only cares about a small number of build constraints (if any), and does not know enough to
specify the rest. For example, a user may know that a certain version of a library like
{\tt boost} is required, but only cares that other build parameters are set so that
the build will succeed.
%
Configuration complexity makes the HPC software ecosystem difficult to manage: there are
simply too many parameters. However, the small set of important build constraints can be
very specific, so we have two competing concerns.  We need the ability to specify details
of the configuration space, without the complexity of remembering all of them.

\subsubsection{Spec Syntax}\label{sec:syntax}

\begin{figure}
\begin{grammar}
  <spec>         ::= <id> [ constraints ]

  <constraints>   ::= \{ `@' <version-list> | `+' <variant> \newline
                   | `-' <variant> ~~~| `~' <variant> \newline
                   | `\%' <compiler> ~| `=' <architecture> \} \newline
                  [ <dep-list> ]

  <dep-list>  ::= \{ `\textsf \textasciicircum' <spec> \}

  <version-list> ::= <version> [ \{ `,' <version> \} ]

  <version>      ::= <id> | <id> `:' | `:' <id> | <id> `:' <id>

  <compiler>     ::= <id> [ <version-list> ]

  <variant>      ::= <id>

  <architecture> ::= <id>

  <id>           ::= [A-Za-z0-9_][A-Za-z0-9_.-]*
\end{grammar}
\caption{
	EBNF grammar for spec expressions.
	\label{fig:grammar}
}
\end{figure}

\begin{table*}\centering
\begin{tabular}{|r|p{2.4in}|p{4in}|}
\hline
& {\bf Spec} & {\bf Meaning} \\
\hline
\hline
1&\small\verb|mpileaks|                         & \small {\tt mpileaks} package, no constraints. \\\hline
2&\small\verb|mpileaks@1.1.2|                   & \small {\tt mpileaks} package, version 1.1.2. \\\hline
3&\small\verb|mpileaks@1.1.2 %gcc|              & \small {\tt mpileaks} package, version 1.1.2, built with {\tt gcc} at the default version. \\\hline
4&\small\verb|mpileaks@1.1.2 %intel@14.1 +debug| & \small {\tt mpileaks} package, version 1.1.2, built with Intel compiler version 14.1, \newline with the ``debug'' build option. \\\hline
5&\small\verb|mpileaks@1.1.2 =bgq|              & \small {\tt mpileaks} package, version 1.1.2, built for the Blue Gene/Q platform (BG/Q). \\\hline
6&\small\verb|mpileaks@1.1.2 ^mvapich2@1.9|     & \small {\tt mpileaks} package version 1.1.2, using {\tt mvapich2}  version 1.9 for MPI. \\\hline
7&\small\verb|mpileaks @1.2:1.4 %gcc@4.7.5 -debug =bgq| \newline
      \verb|  ^callpath @1.1 %gcc@4.7.2| \newline
      \verb|  ^openmpi @1.4.7|                & \small%
      {\tt mpileaks} at any version between 1.2 and 1.4 (inclusive), built with gcc 4.7.5, 
      without the debug option, for BG/Q, linked with {\tt callpath} version 1.1
      and building {\tt callpath} with {\tt gcc} version 4.7.2, linked with {\tt openmpi} version 1.4.7.    \\
\hline
\end{tabular}
\caption{
	Spack build spec syntax examples and their meaning.
	\label{tab:specs}
}
\end{table*}

We have developed a syntax for specs that allows users to specify
only those constraints they care about, but to be specific when necessary.
Our syntax is expressive enough to represent software DAGs  but concise enough to use
on the command line. The spec syntax is recursively defined to allow users to specify
parameters on dependencies as well as on the root of the DAG.  The EBNF grammar we use
to implement this in Spack is shown in Figure~\ref{fig:grammar}.

We begin with an example.%Let us explain our syntax with an example.
Consider the case where a user wants to install the {\tt mpileaks} package, but knows
nothing about its structure.  To install the package, the user writes:
%
\begin{minted}[fontsize=\scriptsize]{bash}
$ spack install mpileaks
\end{minted}
%
Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
Spack parses it and converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
Note that even though the spec contains no dependency information, it is still 
converted to a full DAG, based on the directives supplied in package files. Since there
are no constraints on the nodes, Spack has considerable leeway for how to build the package,
and we say that the package is {\it unconstrained}.
%
Now, suppose the user wants a specific version of {\tt mpileaks}.  This can be requested
with a version constraint after the package name:
%
\begin{minted}[fontsize=\scriptsize]{bash}
$ spack install mpileaks@2.3
\end{minted}
%
We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
If the user does not need a specific version of {\tt mpileaks}, but does require
particular minimu version, then the user could use {\it version range} syntax.
and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save build time if
Spack already has a version installed that satisfies the constraint -- Spack will just take
the already built instead of building a new one.

Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of spec syntax:
%
\begin{minted}[fontsize=\scriptsize]{bash}
$ spack install mpileaks@2.3 ^callpath@1.0+debug ^libelf@0.8.11
\end{minted}
%
The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
we now see that there are now version constraints on {\tt callpath} and {\tt libelf},
and the user has asked for the debug variant of the {\tt callpath} library.

Recall that Spack guarantees there will be only a single version of any package in 
the spec DAG.  Therefore, within the same DAG, each dependency can be uniquely identified by 
only its package name.  The user does not have to think about DAG connectivity to add
constraints.  She need only know that a package depends, somehow on {\tt callpath}. 
For the same reason, the constraint order does not matter; dependency constraints
can appear in arbitrary order.

Table~\ref{tab:specs} shows further examples of specs, ranging from very
simple to more complex. From these examples, we can see that Spack offers constraint
notation to cover the rest of the HPC package parameter space.

{\bf Versions.}
The version constraint, denoted with {\tt @}, was already covered above. Versions
can be precise ({\tt @2.5.1}) or denote a potentially open-ended 
range ({\tt @2.5:}, {\tt @2.5:4.4}).

The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
in our experience users frequently want bleeding-edge versions.  Package managers
frequently lag behind the latest releases. 
Spack has a capability to extrapolate URLs from versions,
using the package's {\tt url} attribute as a model\footnote{This works
for packages with consistently named URLs}.  The user can request a specific
version on the command line, even if it is unknown to Spack,
and Spack will attempt to install it.  Spack also uses the same
model to scrape webpages and find new versions as they become available.

{\bf Compilers.}
With a compiler constraint (shown on line 3) the user
simply adds {\tt \%} followed by its name, along with an optional compiler version 
specifier.  Spack compiler names, e.g. {\tt gcc}, refer to the full compiler {\it toolchain},
i.e. the C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
compiler toolchains if they are in the user's {\tt PATH} or they can be registered manually
through a configuration file.

{\bf Variants.}
To handle build options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementor must check the spec and handle the cases
where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
or {\tt ~debug}).  The names simplify the versioning and prevent
Spack's configuration space from becoming too fine-grained.
It would violate our goal of conciseness, for example, to include detailed 
compiler flags in the spec syntax, but known sets of flags can simply be named.

{\bf Cross-compilation.}
To support cross-compilation, Spack includes the platform in the package spec (line 5).
Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bgq}.  They are
specified per-package; this allows front-end tools to depend on their back-end measurement
libraries with a {\it different} architecture on cross-compiled machines.

\subsubsection{Constraints in packages}

So far, we have shown examples of specs being used to request constraints from the
command line, when {\tt spack install} is invoked.  However, the user is not the only
source of constraints.  Applications may require specific versions of dependencies,
and it is often desirable to write constraints like this into a package file.  For
example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
At first glance, the {\tt depends\_on()} directive in Figure~\ref{fig:mpileaks} look
like they take simple package names.  However, a package name is also a spec, and
the same constraint syntax usable from the command line can be applied inside directives.
So, the ROSE compiler can simply write:
%
\begin{lstlisting}
depends_on(`boost@1.54.0')
\end{lstlisting}
%
This constraint will be incorporated into the initial DAG node generated from
the ROSE package.

Spack also accepts spec syntax for conditionally patching packages.
Spack offers a {\tt patch} directive that uses specs as predicates in an
optional {\tt when} parameter.
These concise directives in the Python package ensure that very specific patches
are only applied to the Python source code when it is being built on Blue Gene/Q,
with the appropriate compiler:
%
\begin{lstlisting}
patch(`python-bgq-xlc.patch',   when=`=bgq%xlc')
patch(`python-bgq-clang.patch', when=`=bgq%clang')
\end{lstlisting}
%
The {\tt when} clause is matched against the Python package's spec
to determine whether the patch will be applied.

Constraint tests can also be used directly within package source code.
For example, in the {\tt install} method, packages can test the versions of dependencies
using constraint syntax:
%
\begin{lstlisting}
  if '' in spec:
  elif '
  '
\end{lstlisting}
%

@@

In some cases, simple conditional logic

In our experience maintaining packages over time, we have found tha

### 2015-04-17 implementation-virtual-deps.tex
Lots of edits.

>>>>

{\bf Dependencies.}
Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
The user can specify all of the above information not only for the package being
installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
\verb|^| and the dependency's name.  If we need to build a new version with a specific
version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
to the spec, and it will build with that MPI version instead of the default.
This can also be done for multiple libraries in the same spec (line 7).

\subsubsection{Querying installed packages}

### 2015-04-17 implementation.tex
GLL edits
Fix typos.  Add gperftools section.
Lots of edits.

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python because it is
very widely used int he HPC community, and because it provides
flexible scripting capabilities. 
%
Like some prior systems, Spack supports arbitrarily many software installations.
It borrows ideas from Nix; packages are installed in their own prefixes,
and configurations are managed using hashes.

\noindent
Spack adds several essential capabilities for HPC:
\begin{enumerate}
\item {\bf Parametric builds} address the combinatorial problem:
      Packages are parameterized by name, version, architecture, compiler, 
      options, and dependencies.
\item To manage the configuration space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
      consistency on the package author.
\end{enumerate}

>>>>

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python for its flexibility
its wide use in the HPC community.
%
Like prior systems, Spack supports an arbitrary number of software
installations, and like Nix it can identify them with hashes.  Unlike any
prior system, Spack provides a language to specify and manage the
combinatorial space of HPC software configurations.

\noindent
Spack has the following unique capabilities:
\begin{enumerate}
\item Packages are {\bf parametrized}, allowing new builds to be composed
      on the fly for different versions, architecture, compiler, options, 
      and dependencies.
\item To manage this large parameter space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that enforce build
      consistency and simplify package writing.
\end{enumerate}

### 2015-04-17 usecase-combinatoric.tex
Fix line wrapping in gperftools
Lots of edits.

Spack presented a solution to both problems.  Site administrators can use Spack to easily maintain a central install of gperftools across the combinations of versions and compilers.  Spack's gperftools package also serves as a central repository for the knowledge of how to build gperftools on each platform and compiler combination.

\begin{figure}
  \includegraphics[width=\columnwidth]{code/gperftools.pdf}
  \caption{
    Simplified install routine for the gperftools tool.
    \label{fig:gperftools}
  }
\end{figure}

>>>>

Spack presented a solution to both problems.  Package administrators can use Spack to 
easily maintain a central install of gperftools across the combinations of versions 
and compilers.  Spack's gperftools package also serves as a central repository for 
the knowledge of how to build gperftools on each platform and compiler combination.

\begin{figure}
\begin{minted}[linenos,
               numbersep=5pt,
			   fontsize=\scriptsize,
               frame=lines,
               framesep=2mm]{python}
patch("patch.gpeftools2.4_xlc", when="@2.4 %xlc")

def install(self, spec, prefix):
    if spec.architecture == "bgq" and self.compiler.satisfies("xlc"):
        configure("--prefix=" + prefix,  "LDFLAGS=-qnostaticlink")
    elif spec.architecture == "bgq":
        configure("--prefix=" + prefix,  "LDFLAGS=-dynamic")
    else:
        configure("--prefix=" + prefix)

    make()
    make("install")
\end{minted}
  \caption{
    Simplified install routine for gperftools.
    \label{fig:gperftools}
  }
\end{figure}
